A tutorial on spatiotemporal partially observed Markov process models via the R package spatPomp

### 2024-10-24 ms.Rnw
getting started on ben bolker suggestions

%% code macros
\newcommand\slot[1]{\code{#1}}
\newcommand\class[1]{class `\code{#1}'}
\newcommand\Class[1]{Class `\code{#1}'}

@@

options(
        scipen=2,
        help_type="html",
        stringsAsFactors=FALSE,
        prompt="R> ",
        continue="+  ",
        width=70,
        useFancyQuotes=FALSE,
        reindent.spaces=2,
        xtable.comment=FALSE
        )

@@

A first step to explore a SpatPOMP model is to simulate stochastic realizations of the latent process and the resulting measurements.
This is carried out by \code{simulate()} which requires specification of \code{rprocess} and \code{rmeasure}.
For example, \code{simulate(bm10)} produces a new object of \class{\spatPomp} for which the original data have been replaced with a simulation from the specified model.
Unless a \code{params} argument is supplied, the simulation will be carried out using the parameter vector in \code{coef(bm10)}.
Optionally, \code{simulate} can be made to return a \class{data.frame} object by supplying the argument \code{format=`data.frame'} in the call to \code{simulate}.

@@

We now move on to specifying our model components as Csnippets.
To get started, we define the movement matrix $\big(v_{\unit,\altUnit}\big)_{\unit,\altUnit \in 1\mycolon\Unit}$ as a global variable in \proglang{C} that will be accessible to all model components, via the \code{globals} argument to \code{spatPomp()}.
<<measles_globals,eval=T,cache=F>>=
measles_globals <- spatPomp_Csnippet("
  const double V[6][6] = {
  {0,2.42,0.950,0.919,0.659,0.786},
  {2.42,0,0.731,0.722,0.412,0.590},
  {0.950,0.731,0,1.229,0.415,0.432},
  {0.919,0.722,1.229,0,0.638,0.708},
  {0.659,0.412,0.415,0.638,0,0.593},
  {0.786,0.590,0.432,0.708,0.593,0}
  }; 
")
@
We now construct a Csnippet for initializing the latent process at time $t_0$.
This is done using unit-specific IVPs, as discussed in Sections~\ref{subsec:init} and~\ref{sec:params}.
Here, the IVPs are \code{S1\_0}$,\dots,$\code{S6\_0}, \code{E1\_0},$\dots$,\code{E6\_0}, and \code{I1\_0},$\dots$,\code{I6\_0}.
These code for the initial value of the corresponding states, \code{S1},$\dots$,\code{S6}, \code{E1},$\dots$,\code{E6}, and \code{I1},$\dots$,\code{I6}.
Additional book-keeping states, \code{C1},$\dots$,\code{C6}, count accumulated cases during an observation interval and so are initialized to zero.
The arguments \code{unit\_ivpnames = c('S','E','I')} and \code{unit\_statenames = c('S','E','I','C')} enable \code{spatPomp()} to expect these variables and define then as needed when compiling the Csnippets.
Similarly, \code{unit\_covarnames = 'P'} declares the corresponding unit-specific population covariate.
This is demonstrated in the following Csnippet specifying \code{rinit}.
<<measles_rinit,echo=T, eval=T,cache=F>>=
measles_rinit <- spatPomp_Csnippet(
  unit_statenames = c('S','E','I','C'),
  unit_ivpnames = c('S','E','I'),
  unit_covarnames = c('P'),
  code = "
    for (int u=0; u<U; u++) {
      S[u] = round(P[u]*S_0[u]);
      E[u] = round(P[u]*E_0[u]);
      I[u] = round(P[u]*I_0[u]);
      C[u] = 0;
    }
  "
)
@
The \code{rprocess} Csnippet has to encode only a rule for a single Euler increment from the process model.
\proglang{C} definitions are provided by \pkg{\spatPomp} for all parameters, state variables, covariates, \code{t}, \code{dt} and \code{U}.
Any additional variables required must be declared as \proglang{C} variables within the Csnippet.
<<measles_rprocess,echo=T, eval=T,cache=F>>=
measles_rprocess <- spatPomp_Csnippet(
  unit_statenames = c('S','E','I','C'),
  unit_covarnames = c('P','lag_birthrate'),
  code = "
    double beta, seas, Ifrac, mu[7], dN[7];
    int u, v;
    int BS=0, SE=1, SD=2, EI=3, ED=4, IR=5, ID=6;

@@

\section*{Acknowledgments}
This work was supported by National Science Foundation grants DMS-1761603 and DMS-1646108, and National Institutes of Health grants 1-U54-GM111274 and 1-U01-GM110712.
We recognize those who have participated in the development and testing of \pkg{spatPomp}, especially Allister Ho, Zhuoxun Jiang, Jifan Li, Patricia Ning, Eduardo Ochoa, Rahul Subramanian and Jesse Wheeler.

>>>>

%% code macros
\newcommand\slot[1]{\code{#1}}
\newcommand\class[1]{class \code{\textquotesingle #1\textquotesingle}}
\newcommand\Class[1]{Class \code{\textquotesingle #1\textquotesingle}}

@@

options(
        scipen=2,
        help_type="html",
        stringsAsFactors=FALSE,
#        prompt="R> ",
#        continue="+  ",
        prompt="   ",
        continue="   ",
        width=70,
        useFancyQuotes=FALSE,
        reindent.spaces=2,
        xtable.comment=FALSE
        )

@@

A first step to explore a SpatPOMP model is to simulate stochastic realizations of the latent process and the resulting measurements.
This is carried out by \code{simulate()} which requires specification of \code{rprocess} and \code{rmeasure}.
For example, \code{simulate(bm10)} produces a new object of \class{\spatPomp} for which the original data have been replaced with a simulation from the specified model.
Unless a \code{params} argument is supplied, the simulation will be carried out using the parameter vector in \code{coef(bm10)}.
Optionally, \code{simulate} can be made to return a \class{data.frame} object by supplying the argument \code{format="data.frame"} in the call to \code{simulate}.

@@

We now move on to specifying our model components as Csnippets.
To get started, we define the movement matrix $\big(v_{\unit,\altUnit}\big)_{\unit,\altUnit \in 1\mycolon\Unit}$ as a global variable in \proglang{C} that will be accessible to all model components, via the \code{globals} argument to \code{spatPomp()}.
<<measles_globals,eval=T,cache=F>>=
measles_globals <- spatPomp_Csnippet("
  const double V[6][6] = {
  {0,2.42,0.950,0.919,0.659,0.786},
  {2.42,0,0.731,0.722,0.412,0.590},
  {0.950,0.731,0,1.229,0.415,0.432},
  {0.919,0.722,1.229,0,0.638,0.708},
  {0.659,0.412,0.415,0.638,0,0.593},
  {0.786,0.590,0.432,0.708,0.593,0}
  }; 
")
@
We now construct a Csnippet for initializing the latent process at time $t_0$.
This is done using unit-specific IVPs, as discussed in Sections~\ref{subsec:init} and~\ref{sec:params}.
Here, the IVPs are \code{S1\_0}$,\dots,$\code{S6\_0}, \code{E1\_0},$\dots$,\code{E6\_0}, and \code{I1\_0},$\dots$,\code{I6\_0}.
These code for the initial value of the corresponding states, \code{S1},$\dots$,\code{S6}, \code{E1},$\dots$,\code{E6}, and \code{I1},$\dots$,\code{I6}.
Additional book-keeping states, \code{C1},$\dots$,\code{C6}, count accumulated cases during an observation interval and so are initialized to zero.
The arguments \code{unit\_ivpnames = c(\textquotesingle S\textquotesingle,\textquotesingle E\textquotesingle, \textquotesingle I\textquotesingle)} and \code{unit\_statenames = c(\textquotesingle S\textquotesingle,\textquotesingle E\textquotesingle,\textquotesingle I\textquotesingle,\textquotesingle C\textquotesingle)} enable \code{spatPomp()} to expect these variables and define then as needed when compiling the Csnippets.
Similarly, \code{unit\_covarnames = \textquotesingle P\textquotesingle} declares the corresponding unit-specific population covariate.
This is demonstrated in the following Csnippet specifying \code{rinit}.
<<measles_rinit,echo=T, eval=T,cache=F>>=
measles_rinit <- spatPomp_Csnippet(
  unit_statenames = c('S','E','I','C'),
  unit_ivpnames = c('S','E','I'),
  unit_covarnames = c('P'),
  code = "
    for (int u=0; u<U; u++) {
      S[u] = round(P[u]*S_0[u]);
      E[u] = round(P[u]*E_0[u]);
      I[u] = round(P[u]*I_0[u]);
      C[u] = 0;
    }
  "
)
@
The \code{rprocess} Csnippet has to encode only a rule for a single Euler increment from the process model.
\proglang{C} definitions are provided by \pkg{\spatPomp} for all parameters, state variables, covariates, \code{t}, \code{dt} and \code{U}.
Any additional variables required must be declared as \proglang{C} variables within the Csnippet.
<<measles_rprocess,echo=T, eval=T,cache=F>>=
measles_rprocess <- spatPomp_Csnippet(
  unit_statenames = c('S','E','I','C'),
  unit_covarnames = c('P','lag_birthrate'),
  code = "
    double beta, seas, Ifrac, mu[7], dN[7];
    int u, v;
    int BS=0, SE=1, SD=2, EI=3, ED=4, IR=5, ID=6;

@@

\section*{Acknowledgments}
This tutorial was supported by National Science Foundation grants DMS-1761603 and DMS-1646108, and National Institutes of Health grants 1-U54-GM111274 and 1-U01-GM110712.
We recognize those who have participated in the development and testing of \pkg{spatPomp}, especially Allister Ho, Zhuoxun Jiang, Jifan Li, Patricia Ning, Eduardo Ochoa, Rahul Subramanian and Jesse Wheeler. Ben Bolker provided feedback incorporated into this version of the article.
